Type-Safe Modular Parsing

### 2017-05-15 SLE_17/sections/RelatedWork.tex
related work

\paragraph{Syntactically Extensible Parsing} Extensible parsing is
used in many different areas. Extensible parser generator is a
mainstream area specially designed for modular syntax and
parsing. Many parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016} support
modular grammars. They allow users to create new modules where new
non-terminals and production rules can be introduced, some can even
override existing rules in the old grammar modules. For instance,
\textit{Rats!}~\cite{Grimm2006} constructs its own module system for
the collection of grammars.  NOA~\cite{Gouseti2014} uses Java
annotation processing to gather information about grammars in Java
code. After all information is collected, NOA merges all grammar
components, produce an ANTLR~\cite{antlr1995} grammar, and generates
corresponding parsing code from the ANTLR grammar. Those parser
generators focus on the \textit{syntactic extensibility} of grammars:
they rely on whole compilation to generate a global parser, even if
there is only a slight modification in the grammar. Some of those
parser generators may statically check the correctness and unambiguity
of grammars. In contrast, because our approach is based on parser
combinators, there is no support for ambiguity checking.  However, as
far as we are aware, no extensible parser generators support separate
compilation or modular type-checking.

Macro systems like the C preprocessor, C++ templates and
Racket~\cite{Tobin-Hochstadt2011}, and other meta-programming
techniques are a similar area aiming at syntactic extensibility.
SugarJ~\cite{Erdweg2011} conveniently introduces syntactic sugar for
Java using library imports. Composition of syntactic sugar is easy for
users, but it requires many rounds of parsing and adaption. This
significantly affects efficiency of compilation. Since the
implementation was based on SDF~\cite{Heering1989} and
Stratego~\cite{Visser2001}, it does not support separate
compilation. Racket adopts a macro system for library-based language
extensibility~\cite{Tobin-Hochstadt2011}. This is quite powerful and it uses
attributed ASTs for contextual
information. Moreover extensions can be integrated in a modular
way. However such modularity is not flexible enough for language
unification, as the syntax is only built from extensions.
%Moreover,
%existing macros cannot be further changed after
%definition. \haoyuan{correct?}
Extensible
compilers like JastAdd~\cite{Ekman2007} and
Polyglot~\cite{Nystrom2003} also support extensible parsing, but this
is mostly done using parser generators. They focus on the
extensions to a host language. Our technique is designed for type-safe
modular language parsers (not just language
extensions), with very flexible parser composition.

@@

\paragraph{Extensible Parsing Algorithm}
Extensible parsing algorithm is another area that relates to
separate compilation.
\textit{Parse table composition}~\cite{bravenboer2008parse}
is an approach where grammars are compiled to
modular parse tables. Those parse tables are expressed as DFAs
or NFAs, and later they can be composed by an algorithm, to provide
separate compilation for parsing. The generation of parse tables can
be quite expensive in terms of performance. The approach
is quite different from ours, since it uses parse
 tables, whereas we use parser combinators.
Our approach supports both
\emph{separate compilation} as well as \emph{modular
  type-checking}. Moreover, the extensibility of parsing is further
available at language composition and lexical level.

@@

\paragraph{Parser Combinators} Since Burge~\cite{burge1975} firstly
introduced parser generators, and after Wadler popularized them
in functional programming~\cite{Wadler1985}, parser combinators have been
more and more popularly developed and used in the research area of
parsing. Many parsing libraries produce recursive descent
parsers by introducing functional monadic
parser combinators~\cite{nott237}. Parsec~\cite{Leijen2001} is
perhaps the most popular parser combinator library in this line.
It is widely used in Haskell (with various ``clones'' in other languages)
for context-sensitive grammars with infinite lookahead. Nevertheless,
Parsec users suffer from manual left-recursion elimination,
high cost for backtracking and longest match composition issues,
as we discussed in Section~\ref{subsec:challenges}. Those limitations make Parsec
(and similar parsing techniques) inadequate for modular parsing.

Some more recent work on parser
combinators~\cite{Ford2002,Might2011,Frost2008} proposed a series of
novel parsing techniques that address the issue of
left-recursion. We have selected Packrat parsing for our approach,
as the combinators are convenient to use in Scala, and
integrate well with multiple trait inheritance and
Object Algebras.
%Also,
%its support for direct left-recursion is already helpful for designing
%real-world parsers practically, though~\cite{warth2008} further
%demonstrated that general left-recursion is supported theoretically.
Despite our choice of Packrat parsing, as argued in Section~\ref{sec:packrat},
any parser combinator approach which supports left-recursion, longest
match composition, and makes the cost of backtracking reasonable,
should be applicable to modular parsing.
%Yet we have mentioned that the components used in our approach can
%have alternatives to achieve better modular parsing.

\paragraph{Extensibility} Various design patterns~\cite{gamma1995design} in multiple
languages, have been proposed over the years to address extensibility
problems, such as the Expression Problem~\cite{wadler1998expression}. However, to the best
of our knowledge, none of the existing extensibility techniques
address the problem of \emph{modularly
parsing extensible ASTs}.  There has been
a lot of work on extensible ASTs, in both functional and
object-oriented programming.  For instance, the famous ``Datatypes \`a
la Carte'' (DTC)~\cite{swierstra2008data} approach represents modular ASTs using co-products
of every two functors, which forms a binary tree structure.
Several variants of DTC have been later proposed~\cite{Bahr2011,Bahr2014,Oliveira2015}.
All of that work essentially covers how to traverse and
consume extensible ASTs.
Only in Bahr's~\cite{Bahr2011} work a brief mention of \emph{unfolds}
(which are a class of algorithms that \emph{build} ASTs) are mentioned.
However Bahr's work does not cover parsing.

In OO languages, there are also many design patterns that achieve
type-safe extensibility~\cite{torgersen2004expression,odersky2005independently,oliveira2009modular,Oliveira:2012,wang2016expression}. We chose Object Algebras~\cite{Oliveira:2012} because the pattern is
relatively lightweight and makes good use of existing OO features,
such as inheritance, generics and subtyping. As seen throughout the paper,
the parsing code is concise and expressive using Object Algebras.
In principle our approach is not coupled to the use of Object
Algebras, and variants of our approach could be used with other
extensibility techniques. Similarly to the work in functional
programming, the extensibility techniques in OOP have focused
on traversing and processing ASTs. We are unware of any work on OOP
that covers how to do modular parsing for extensible ASTs.

It is worth mentioning that Scala case classes~\cite{emir2007matching} provide a near
solution to the Expression Problem. Case classes are similar to
algebraic datatypes but, unlike algebraic datatypes, new cases can be
modularly added. The reason why case classes are not a complete
solution to the Expression Problem is that they do not enforce
exhaustiveness of pattern matching for extensible operations. In other
words run-time pattern matching errors can happen when writing extensible code with case classes. So that full static type-safety is not ensured. Nevertheless case classes are a pragmatic approach, which is
widely used in practice. Therefore, modular parsing techniques may be
of value for extensible code using case classes.  The approach we
presented in Section~\ref{sec:inheritance} can readily be adapted to case classes:
all that the users need to do is to use case classes, instead of
standard OO classes in their code.

>>>>

\paragraph{Syntactically Extensible Parsing}
Extensible parser generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016}
are a mainstream area of modular syntax and parsing. They allow users to write
modular grammars, where new
non-terminals and production rules can be introduced, some can even
override existing rules in the old grammar modules. For instance,
\textit{Rats!}~\cite{Grimm2006} constructs its own module system for
the collection of grammars, while NOA~\cite{Gouseti2014} uses Java
annotation to collect all information before producing an ANTLR~\cite{antlr1995} grammar
and the parsing code. Those parser
generators focus on the \textit{syntactic extensibility} of grammars:
they rely on whole compilation to generate a global parser, even if
there is only a slight modification in the grammar. Some of those
parser generators may statically check the correctness and unambiguity
of grammars. In contrast, because our approach is based on parser
combinators, there is no support for ambiguity checking.  However, as
far as we are aware, no extensible parser generators support separate
compilation or modular type-checking.

Macro systems like the C preprocessor, C++ templates and
Racket~\cite{Tobin-Hochstadt2011}, and other meta-programming
techniques are a similar area aiming at syntactic extensibility.
SugarJ~\cite{Erdweg2011} conveniently introduces syntactic sugar for
Java using library imports. Composition of syntactic sugar is easy for
users, but it requires many rounds of parsing and adaption, hence
significantly affects the efficiency of compilation. Since the
implementation was based on SDF~\cite{Heering1989} and
Stratego~\cite{Visser2001}, it does not support separate
compilation. Racket adopts a macro system for library-based language
extensibility~\cite{Tobin-Hochstadt2011}. It uses
attributed ASTs for contextual
information, and extensions can be integrated in a modular
way. However such modularity is not flexible enough for language
unification, as the syntax is only built from extensions.
%Moreover,
%existing macros cannot be further changed after
%definition. \haoyuan{correct?}
Extensible
compilers like JastAdd~\cite{Ekman2007} and
Polyglot~\cite{Nystrom2003} also support extensible parsing, but this
is mostly done using parser generators. They focus on the
extensions to a host language. Those techniques are short of type safety in a modular
setting as well.

@@

\paragraph{Extensible Parsing Algorithm}
\textit{Parse table composition}~\cite{bravenboer2008parse}
is an approach where grammars are compiled to
modular parse tables. Those parse tables are expressed as DFAs
or NFAs, and later they can be composed by an algorithm, to provide
separate compilation for parsing. The generation of parse tables can
be quite expensive in terms of performance. The approach
is quite different from ours, since it uses parse
 tables, whereas we use parser combinators.
Our approach supports both
\emph{separate compilation} as well as \emph{modular
  type-checking}. Moreover, the extensibility of parsing is further
available at language composition and lexical level.

@@

\paragraph{Parser Combinators} Parser combinators have become more and more
popular since~\cite{burge1975,Wadler1985}. Many parsing libraries produce recursive descent
parsers by introducing functional monadic
parser combinators~\cite{nott237}. Parsec~\cite{Leijen2001} is
perhaps the most popular parser combinator library in this line.
It is widely used in Haskell (with various ``clones'' in other languages)
for context-sensitive grammars with infinite lookahead. Nevertheless,
Parsec users suffer from manual left-recursion elimination,
high cost for backtracking and longest match composition issues,
as we discussed in Section~\ref{subsec:challenges}. Those limitations make Parsec
(and similar parsing techniques) inadequate for modular parsing.

Some more recent work on parser
combinators~\cite{Ford2002,Might2011,Frost2008} proposed a series of
novel parsing techniques that address the issue of
left-recursion. We have selected Packrat parsing due to its simplicity in Scala,
but in general it can also have alternatives.

\paragraph{Extensibility} Various design patterns~\cite{gamma1995design} in multiple
languages, have been proposed over the years to address extensibility
problems, such as the Expression Problem~\cite{wadler1998expression}.
The famous ``Datatypes \`a
la Carte'' (DTC)~\cite{swierstra2008data} approach represents modular ASTs using co-products
of every two functors. Several variants of DTC have been later proposed~\cite{Bahr2011,Bahr2014,Oliveira2015}.
All of that work essentially covers how to traverse and
consume extensible ASTs. However they do not
address the problem of \emph{modularly
parsing extensible ASTs}. Only in Bahr's~\cite{Bahr2011} work \emph{unfolds} is briefly mentioned,
yet it does not cover parsing.

There are also many design patterns in OO languages that achieve
type-safe extensibility~\cite{torgersen2004expression,odersky2005independently,oliveira2009modular,Oliveira:2012,wang2016expression}. We chose Object Algebras~\cite{Oliveira:2012} because the pattern is
relatively lightweight and makes good use of existing OO features,
such as inheritance, generics and subtyping. As seen throughout the paper,
the parsing code is concise and expressive using Object Algebras.
On the other hand we are unware of any work on OOP
that has covered how to do modular parsing for extensible ASTs.

It is worth mentioning that Scala case classes~\cite{emir2007matching} provide a near
solution to the Expression Problem. Case classes can be
modularly added, but they do not enforce
exhaustiveness of pattern matching for extensible operations. In other
words run-time pattern matching errors can happen when writing extensible code with case classes. So that full static type-safety is not ensured. Nevertheless case classes are a pragmatic approach, which is
widely used in practice. Therefore, modular parsing techniques may be
of value for extensible code using case classes.  The approach we
presented in Section~\ref{sec:inheritance} can readily be adapted to case classes:
all that the users need to do is to use case classes, instead of
standard OO classes in their code.\haoyuan{To simplify; conclusion}
